---
title: " Interval censored Regression Tree using Survreg"
author: "Parismita Das"
date: "11 January 2018"
subtitle: "Hard"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To implement a (slow) regression tree for interval censored output data, using the survival::survreg function, which can be used to fit a linear model for censored outputs. Then search for the best possible split over all features by maximizing the logLik of the survreg model. 

```{r m1, echo=FALSE}
library(partykit)
library(survival)
set.seed(290875)
```

#Get Best Split

Function get_split splits and checks for the best split by maximizing the loglik of survreg model.
We surf through all possible splits by using 2 nested loops on rows and columns. Split and every point and check if that is the maximum or not.

As this process takes a lot of time, as it has a Time complexity of O(n^2) Hence its known as Slow Regression.

```{r m2, echo=TRUE}
##########################split index##############################################
get_split <- function(y, dataset,i,min_size){
  #split object gives info about the best split
  split<-NULL
  split$log<- -Inf
  split$group$right<-NULL
  split$group$left<-NULL
  split$index <- 0
  split$row <-0
  split$br <- 0
  
  for (index in 1:(length(dataset[0,]))){ #4
    if(i!=index){
      
      #the formula used in survreg for fitting
      form <-as.formula(paste(y, 
                              paste(names(dataset[index]), 
                                    collapse="+"),collapse = "+"))
      
      for (row in 1:length(dataset[,index])){ #150
        br<-dataset[row,index]
        group <- test_split(dataset, index, br)
        ctrl<-survreg.control(maxiter=60, rel.tolerance=1e-06,toler.chol=1e-7)
        if((length(group$left)>min_size)&&(length(group$right)>min_size))
        {
          #fitting left and right part of split and summing their loglikelihood
          fitl <- survreg(form, dataset[group$left,] ,dist = "t",control = ctrl)
          l<-logLik(fitl)
          fitr <- survreg(form, dataset[group$right,] ,dist = "t",control = ctrl)
          r<-logLik(fitr)
          log<- l+r
          
          #checking condition if loglikelihood is Inf or Nan or 0 
          #then it should be excluded.
          if(is.nan(log)||is.nan(r)||is.nan(l)||(log==0)){
            log<--Inf
          }
          
          #maximizing the loglikelihood
          if (split$log < as.numeric(log)){
            split$index <- index
            split$row <-row
            split$br <- br
            split$log <- log
            split$group<-group
            
          }
        }
      }
    }
  }
  return(split)
}
```

#Split The Dataset

This function splits the dataset into 2 parts : left and right by checking if value is lesser or higher.
For lesser value than break point, the datapoint goes to left side and vise-versa.
```{r m3, echo=TRUE}
####################right-left split#################################
test_split <- function(dataset, index, value){
  left <- c()
  right <- c()
  for (row in 1:length(dataset[,2])){
    if (as.numeric(dataset[row,index]) < value){
      left <- cbind(left,row)
    }
    else{
      right <- cbind(right,row)
    }
  }
  group <-c()
  group$left <- left
  group$right <- right
 
  return (group)
}
```

#Structure The Tree

The split function splits the dataset according to the best split point we get from get_split. We use depth first search to split the tree and  make a binary tree out of it. 

Note: Here we consider only numerical columns for splitting.

```{r m4, echo=TRUE}
############################3tree######################################
# Create child splits for a node or make terminal
split <- function(y,sp, depth, maxdepth,min_size,i,dataset,b_index){
  rsize<- length(sp$group$right)
  lsize<- length(sp$group$left)
  #b_index
  l<- length(b_index[,1])
  id <- as.numeric(b_index[l,1])+1
  term <- 0
  varid<-0
  kidi<-0
  kidf<-0
  br<-0
  info <- ''
  tag<-'no'
  
  # check for a no split
  if (!((any(sp$group$left))&&(any(sp$group$right)))){
    return(b_index)
  }
  
  #splitting into right
  tag<-'right'
  #print("right")
  #new splitiing object
  spn<- get_split(y,dataset[sp$group$right,],i,min_size)
  
  #if its a terminal
  if((spn$log==-Inf)||(spn$log==0)||(as.numeric(depth)>as.numeric(maxdepth))
     ||(as.numeric(rsize)<as.numeric(min_size))){
    term <- 1
    sp$group$right <- NULL
    node <- c(id,varid,br,term,kidi,kidf,info,tag)
    b_index <- rbind(b_index,node)
  }
  #if we require further right splitting
  else{
    varid <- spn$index
    node <- c(id,varid,spn$br,term,kidi,kidf,info,tag)
    b_index <- rbind(b_index,node)
    b_index <- split(y,spn,depth+1,maxdepth,
                     min_size,i,dataset[sp$group$right,],b_index)
  }
  
  #b_index
  tag<-'left'
  #print("left")
  l<- length(b_index[,1])
  id <- as.numeric(b_index[l,1])+1
  term <- 0
  varid<-0
  br<-0
  info <- ''
  
  #splitting into left
  spn <- get_split(y,dataset[sp$group$left,],i,min_size)
  
  #terminal
  if((spn$log==-Inf)||(spn$log==0)||(as.numeric(depth)>as.numeric(maxdepth))
     ||(as.numeric(lsize)<as.numeric(min_size))){
    term <- 1
    sp$group$left <- FALSE
    node <- c(id,varid,br,term,kidi,kidf,info,tag)
    b_index <- rbind(b_index,node)
  }
  #splitting further
  else{
    varid <- spn$index
    br<- spn$br
    node <- c(id,varid,br,term,kidi,kidf,info,tag)
    b_index <- rbind(b_index,node)
    gr <- spn$group
    b_index <- split(y,gr,depth+1,maxdepth,min_size,i,
                     dataset[sp$group$left,],b_index)
  }
  
  return(b_index)
}

```

#Kids Index

We Use the same algo here as in Medium Test.

```{r m6, echo=TRUE}
###############################terminal output############################
kid <- function(b_index){
  n<- 0
  k<-c()
  
  for(i in 2:(length(b_index[,1]))){
    #if tag is right and next tag is left -> put into i-1th b_index
    if(b_index[i,8]=="right"){
      b_index[i-1,6] = i
      if(b_index[i+1,8]=="left"){
        b_index[i-1,5] = i+1      
      }
    }
    #check of left out left tags
    if((b_index[i,8]=="left")&&(b_index[i-2,5]==0)){
      k<-c(k,i)
    }
  }
  c<-1
  #put k from botton to top
  for(i in length(b_index[,1]):1){
    if((b_index[i,5]==0)&&(b_index[i,4]==0)&&(b_index[i,6]!=0)){
      b_index[i,5]=k[c]
      c=c+1
    }
  }
  
  #If kidi and kidf are 0 and terminal is 0 then make it 1 
  for(i in 2:length(b_index[,1])){
    if((b_index[i,5]==0)&&(b_index[i,5]==0)){
      b_index[i,4]=1
    }
  }
  return (b_index)
}

```

#Partysplit and Partynode

We are making list of partysplit objects and creating partynode which is used to create The Tree.

```{r m7, echo=TRUE}
################################cart algo##################################
cart <- function(y,dataset,i,maxdepth,min_size){
  b_index <- c(1)
  
  #defining formula and dataset
  form <-as.formula(paste(y, "."))
  dataset <- model.frame(form, data = dataset)

  #root splitting
  b <- get_split(y,dataset,i,min_size)
  
  #defininf b_index
  b_index <- data.frame(b_index,b$index,b$br,0,0,0,'',
                        "root",stringsAsFactors=FALSE)
  colnames(b_index) <- c("id","varid","break","terminal",
                         "kidi","kidf","info","tag")
  
  #further splitting
  b_index <- split(y,b,1,maxdepth,min_size,i,dataset,b_index)
  b_index <- kid(b_index)
  print(b_index)

  #list of partysplit objects
  nodelist <- list(
    #root node
    list(id = 1L, split = partysplit(varid = as.integer(b_index[1,2]), 
                                     breaks = as.numeric(b_index[1,3])), 
         kids = c(as.integer(b_index[1,5]),as.integer(b_index[1,6]))))
  
  for(i in 2:length(b_index[,1])){
    #terminal
    if(as.numeric(b_index[i,4])){
      nodelist[i] <- list(list(id = as.integer(b_index[i,1]), info = b_index[i,7]))
    }
    #mid-values
    else{
      nodelist[i] <- list(
        list(id = as.integer(b_index[i,1]), split = partysplit(
          varid = as.integer(b_index[i,2]), 
          breaks = as.numeric(b_index[i,3])), 
          kids = c(as.integer(b_index[i,5]),as.integer(b_index[i,6])),
          info = b_index[i,7]))
    }
  }
  ## convert to a recursive structure
  node <- as.partynode(nodelist)
  
  ## set up party object
  tree <- party(node, data = dataset, fitted = data.frame(
    "(fitted)" = fitted_node(node, data = dataset),
    "(response)" = model.response(dataset),
    check.names = FALSE), terms = terms(dataset))
  tree <- as.constparty(tree)
  
  return(tree)
}
```

# The Result

Here is the b_index value and the generated Tree. 
I am taking 45 datapoints of the neuroblastomaProcessed dataset and 118 features.

```{r m8, warning=F, echo=FALSE}
#########################main code#####################################
data(neuroblastomaProcessed, package="penaltyLearning")

survTarget <- with(neuroblastomaProcessed, {Surv(target.mat[, 1], 
                                                 target.mat[,2], type="interval2")})
trainData <- data.frame(log.penalty=survTarget, neuroblastomaProcessed$feature.mat)


y <- "log.penalty~"
form <-as.formula(paste(y, paste(names(trainData[2]), collapse="+"),collapse = "+"))
fit <- survreg(form, trainData,dist = "t")
l<-logLik(fit)

print("b_index:")
fit<- cart(y,trainData[1:45,],1,4,2)
print("The tree:")
plot(fit)
```

